
# Class Activity. Introduction to R, Part 1. September 1, 2022

## Reading in Data

```{r}
library(tidyverse)
library(here)
```

```{r}
county_elections <- read_csv(here("data", "county_elections.csv"))

# Remove any rows with missing values to avoid warning messages
county_elections <- na.omit(county_elections)
```






## Visualization

### Univariate Graphs

```{r, warning=FALSE, message=FALSE}
ggplot(county_elections) +
  aes(x = median_hh_inc) +
  geom_histogram()
```



```{r, warning=FALSE}
ggplot(county_elections) +
  aes(x = median_hh_inc) +
  geom_histogram(bins = 50,
                 color = "white",
                 fill = "steelblue")
```

```{r, warning=FALSE}
ggplot(county_elections) +
  aes(x = median_hh_inc) +
  geom_histogram(bins = 50,
                 color = "white",
                 fill = "steelblue") +
  labs(title = "Distribution of Median County Incomes",
       x = "Median Household Income",
       y = "Count")
```

```{r, warning=FALSE}
ggplot(county_elections) +
  aes(x = median_hh_inc) +
  geom_histogram(bins = 50,
                 color = "white",
                 fill = "steelblue") +
  labs(title = "Distribution of Median County Incomes",
       x = "Median Household Income",
       y = "Count") +
  theme_minimal()
```

```{r, warning=FALSE}
ggplot(county_elections) +
  aes(x = median_hh_inc) +
  geom_histogram(bins = 50,
                 color = "white",
                 fill = "steelblue") +
  labs(title = "Distribution of Median County Incomes",
       x = "Median Household Income",
       y = "",
       caption = "(log10 scale)") +
  theme_minimal() +
  scale_x_log10(labels = scales::dollar)
```

*Class Activity*:

Create a univariate graph of your choosing using ggplot. Try other types of geoms besides histograms.


```{r}
ggplot(county_elections) +
  aes(x = clf_unemploy_pct) +
  geom_density()
```



### Bivariate Graphs

```{r, warning=FALSE}
ggplot(county_elections) +
  aes(x = median_hh_inc, y = lesshs_pct) +
  geom_point()
```

```{r, warning=FALSE}
ggplot(county_elections) +
  aes(x = median_hh_inc, y = lesshs_pct) +
  geom_point() +
  labs(title = "US Counties by Education and Income",
       x = "Median Household Income",
       y = "Less than High School %") +
  scale_x_continuous(label = scales::dollar) +
  theme_classic()
```

```{r, warning=FALSE}
ggplot(county_elections) +
  aes(x = median_hh_inc, y = lesshs_pct) +
  geom_point(alpha = 0.2, color = "darkcyan") +
  labs(title = "US Counties by Education and Income",
       x = "Median Household Income",
       y = "Less than High School %") +
  scale_x_continuous(label = scales::dollar) +
  theme_classic()
```

```{r, warning=FALSE}
ggplot(county_elections) +
  aes(x = median_hh_inc, y = lesshs_pct) +
  geom_point(alpha = 0.2, color = "darkcyan") +
  geom_smooth(method = "lm", color = "black") +
  labs(title = "US Counties by Education and Income",
       x = "Median Household Income",
       y = "Less than High School %") +
  scale_x_continuous(label = scales::dollar) +
  theme_classic()
```

```{r, warning=FALSE}
ggplot(county_elections) +
  aes(x = median_hh_inc, y = lesshs_pct) +
  geom_point(alpha = 0.2, color = "darkcyan") +
  geom_smooth(method = "lm", color = "black",
              formula = "y ~ log(x)") +
  labs(title = "US Counties by Education and Income",
       x = "Median Household Income",
       y = "Less than High School %") +
  scale_x_continuous(label = scales::dollar) +
  theme_classic()
```

*Class Activity*:

Create a bivariate graph of your choosing using ggplot



### Trivariate(!) Graphs

```{r}
county_elections <- county_elections |> 
  mutate(rural = ifelse(rural_pct > 50, "Rural", "Not Rural"))
```

```{r, warning=FALSE}
ggplot(county_elections) +
  aes(x = median_hh_inc, y = lesshs_pct,
      color = rural) + # Coloring points based on rural variable
  geom_point(alpha = 0.5) + # Removed color from geom
  labs(title = "US Counties by Education and Income",
       x = "Median Household Income",
       y = "Less than High School %") +
  scale_x_continuous(label = scales::dollar) +
  theme_classic()
```

```{r, warning=FALSE}
ggplot(county_elections) +
  aes(x = median_hh_inc, y = lesshs_pct,
      size = total_population) + # Changing size of points
  geom_point(alpha = 0.2) + 
  labs(title = "US Counties by Education and Income",
       x = "Median Household Income",
       y = "Less than High School %") +
  scale_x_continuous(label = scales::dollar) +
  theme_classic()
```

```{r}
ggplot(county_elections) +
  aes(x = median_hh_inc, y = lesshs_pct) +
  geom_point(alpha = 0.2, color = "darkcyan") +
  labs(title = "US Counties by Education and Income",
       x = "Median Household Income",
       y = "Less than High School %") +
  scale_x_continuous(label = scales::dollar) +
  theme_classic() +
  facet_wrap(~ rural) # Adding faceting
```

```{r}
ggplot(county_elections) +
  aes(x = median_hh_inc, y = lesshs_pct) +
  geom_point(alpha = 0.2, color = "darkcyan") +
  labs(title = "US Counties by Education and Income",
       x = "Median Household Income",
       y = "Less than High School %") +
  scale_x_continuous(label = scales::dollar) +
  theme_classic() +
  facet_wrap(~ rural, nrow = 2)
```

*Class Activity*:

Using either colors, sizes, or facets, modify your bivariate graph to include information from a third variable.



### Choropleth Maps

```{r}
library(urbnmapr)
```

```{r}
county_elections <- county_elections |> 
  mutate(county_fips = str_pad(countyCode, width = 5, pad = "0"))
```

```{r}
map_data <- left_join(county_elections, counties,
                      by = "county_fips")
```

```{r}
ggplot(map_data) +
  aes(x = long, y = lat, 
      group = group, fill = lesshs_pct) +
  geom_polygon(color = NA) +
  # This second geom_polygon shows the state borders
  geom_polygon(data = states, mapping = aes(long, lat, group = group),
               fill = NA, size = 0.1, color = "white") +
  # Making maps requires you to choose a geographic projection
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  # theme_void gives us a blank canvas
  theme_void()
```

```{r}
ggplot(map_data) +
  aes(x = long, y = lat, 
      group = group, fill = lesshs_pct) +
  geom_polygon(color = NA) +
  geom_polygon(data = states, mapping = aes(long, lat, group = group),
               fill = NA, size = 0.1, color = "white") +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  # This creates a diverging color scale
  # that is also colorblind friendly
  scale_fill_viridis_c() +
  labs(fill = "Less than High School %") +
  theme_void() +
  theme(legend.position = "bottom") 
```

*Class Activity*:

Choose a new variable from `county_elections` and create a choropleth map of it by adjusting the code above