
# Class Activity. Introduction to R, Part 2. September 6, 2022

Chapters 3 and 4 in https://ucsdpolimathcamp.github.io/MathCamp/

## Data Cleaning

```{r}
# run in your Console

# install.packages("unvotes")
```

```{r, message=FALSE, warning=FALSE}
library(unvotes)
library(tidyverse)
library(here)
```

```{r}
un_votes <- un_votes
un_roll_calls <- un_roll_calls
un_roll_call_issues <- un_roll_call_issues
```

About the data

-   `un_votes`, country-vote level data. Each row is a country's vote on a particular UN Assembly resolution.
-   `un_roll_calls`, resolution level data. Contains information about each resolution.
-   `un_roll_call_issues`, resolution level data. Contains the issue-area for each resolution.

### Pipe Operator

```{r, eval=FALSE}
# The pipe way
me |> 
  wake_up(time = "8:00") |> 
  get_out_of_bed(side = "correct") |> 
  get_dressed(pants = TRUE, shirt = TRUE) |> 
  leave_house(car = FALSE, bike = TRUE)

# The non-pipe way
leave_house(get_dressed(get_out_of_bed(wake_up(me, time = "8:00"), side = "correct"), pants = TRUE, shirt = TRUE), car = FALSE, bike = TRUE)
```


### Select

```{r}
un_roll_calls |> 
  select(rcid, date) # Keeping only two variables
```

```{r}
# This creates a new object called "small_un_roll_calls"
# containing only rcid and date columns
small_un_roll_calls <- un_roll_calls |> 
  select(rcid, date)
```

```{r}
un_roll_calls |> 
  select(-session) # Keep everything except session column
```

```{r}
un_roll_calls |> 
  select(where(is.character)) # Keep all character columns
```

```{r}
un_roll_calls |> 
  select(un_resolution = unres, date) # Select and rename columns
```

Renaming without selecting: 

```{r}
un_roll_calls |> 
  rename(un_resolution = unres,
         amendment = amend,
         paragraph = para)
```

### Mutate

```{r}
un_votes |> 
  mutate(rcid_10 = rcid * 10) # Creates the new rcid_10 variable
```

```{r}
un_votes |> 
  mutate(rcid = rcid * 10) # Overwrites the existing rcid variable
```

```{r}
un_votes |> 
  mutate(vote_dummy = case_when(vote == "yes" ~ 1,
                                vote == "no" ~ 0))
```

```{r}
un_votes |> 
  mutate(rcid_era = case_when(rcid < 2000 ~ "old",
                              rcid >= 2000 & rcid < 6000 ~ "middle",
                              rcid >= 6000 ~ "recent"))
```

To recap, the most commonly used functions for cleaning columns/variables in your data are:

  -   `select()` removes and/or renames columns.
  -   `rename()` renames existing columns without removing any.
  -   `mutate()` changes the values of existing columns and creates new columns.


### Filter

```{r}
un_roll_calls |> 
  filter(importantvote == 1)
```

```{r}
# & for AND
un_roll_calls |> 
  filter(importantvote == 1 & session > 40)

# | for OR
un_roll_calls |> 
  filter(importantvote == 1 | session > 40)
```

```{r}
un_roll_calls |> 
  filter(!is.na(amend)) # Removing rows with NA for the amend variable
```

```{r}
# Cumbersome way
un_votes |> 
  filter(country == "Kenya" |
         country == "Grenada" |
         country == "Canada" |
         country == "Latvia" |
         country == "Yemen" |
         country == "Angola")

# Easier way
country_list <- c("Kenya", "Grenada", "Canada",
                  "Latvia", "Yemen", "Angola")
un_votes |> 
  filter(country %in% country_list)
```

### Aggregation

```{r}
un_votes |> 
  # Creating a new numeric dummy vote variable
  mutate(vote_dummy = case_when(vote == "yes" ~ 1,
                                vote == "no" ~ 0)) |> 
  # Specify level of aggregation
  group_by(country) |> 
  # Perform the aggregation function by group
  summarize(proportion_yes_vote = mean(vote_dummy, na.rm = TRUE))
```

```{r}
un_votes |> 
  mutate(vote_dummy = case_when(vote == "yes" ~ 1,
                                vote == "no" ~ 0)) |> 
  group_by(country) |> 
  summarize(proportion_yes_vote = mean(vote_dummy, na.rm = TRUE)) |> 
  arrange(proportion_yes_vote)

# Use arrange(desc(proportion_yes_vote))
# to sort in descending order
```

```{r}
un_roll_call_issues |> 
  count(issue)
```

### Merging Data

```{r, eval=FALSE}
xy_data <- left_join(x, y, by = "key_variable")
```

```{r, eval=FALSE}
xy_data <- x |> 
  left_join(y, by = "key_variable")
```

```{r}
un_votes |> 
  left_join(un_roll_calls, by = "rcid")
```

https://github.com/gadenbuie/tidyexplain




## Loops and Simulations

### sample()

```{r}
# Create a vector of character values
my_animals <- c("cat", "dog", "fish", "monkey", "parrot")

# From the set "my_animals", draw 2 at random
sample(x = my_animals, size = 2)
```

```{r, error=TRUE}
sample(x = my_animals, size = 100)
```

We set `error=TRUE` in the code chunk above to tell R we expect the code to make an error. This allows us to knit the RMarkdown document despite the coding error.

```{r}
sample(x = my_animals, size = 100, replace = TRUE)
```

```{r}
my_animals <- sample(
  x = my_animals,
  size = 1000,
  replace = TRUE,
  prob = c(0.1, 0.2, 0.2, 0.4, 0.1)
)
```

```{r}
# ggplot requires the data be in tibble or data.frame form
tibble(my_animals) |> 
  ggplot(aes(x = my_animals)) +
  geom_bar()
```

### Sampling from Distributions

### Binomial Distribution

```{r}
rbinom(n = 100, size = 1, prob = 0.5)
```

```{r}
# This does the same thing as the chunk above
sample(x = c(0, 1),
       size = 100,
       replace = TRUE)
```

```{r}
rbinom(n = 100, size = 1, prob = 0.8)
```

```{r}
binomial_draws <- rbinom(n = 100, size = 1, prob = 0.8)

tibble(binomial_draws) |> 
  ggplot(aes(x = binomial_draws)) +
  geom_bar() +
  scale_x_continuous(breaks = 0:1) +
  labs(title = "Binomial Distribution: size = 1, prob = 0.8")
```

```{r}
rbinom(n = 1, size = 100, prob = 0.5)
```

```{r}
binomial_draws <- rbinom(n = 10000,
                         size = 7,
                         prob = 0.5)
```

```{r}
tibble(binomial_draws) |> 
  ggplot(aes(x = binomial_draws)) +
  geom_bar() +
  scale_x_continuous(breaks = 0:7) +
  labs(title = "Binomial Distribution: size = 7, prob = 0.5")
```

```{r}
binomial_draws <- rbinom(n = 10000,
                         size = 7,
                         prob = 0.8)

tibble(binomial_draws) |> 
  ggplot(aes(x = binomial_draws)) +
  geom_bar() +
  scale_x_continuous(breaks = 0:7) +
  labs(title = "Binomial Distribution: size = 7, prob = 0.8")
```

### Uniform Distribution

```{r}
# Generating random numbers from between 0 and 1
runif(n = 25, min = 0, max = 1)
```

```{r, waring=FALSE}
uniform_draws <- runif(10000, min = 0, max = 1)

tibble(uniform_draws) |> 
  ggplot(aes(x = uniform_draws)) +
  geom_histogram(color = "white", bins = 30) +
  xlim(-0.5, 1.5) +
  labs(title = "Unfiform Distribution: min = 0, max = 1")
```

### Normal Distribution

```{r}
rnorm(n = 100, mean = 0, sd = 1)
```

```{r}
normal_draws <- rnorm(10000, mean = 0, sd = 1)

tibble(normal_draws) |> 
  ggplot(aes(x = normal_draws)) +
  geom_histogram(color = "white", bins = 30) +
  labs(title = "Normal Distribution: mean = 0, sd = 1")
```

```{r}
# rnorm defaults to mean = 0, sd = 1 if those options are left blank
normal_minus_one <- rnorm(10000, mean = -1)
normal_zero <- rnorm(10000)
normal_one <- rnorm(10000, mean = 1)

# Combining the draws into a single "long" data frame
normal_data <- cbind(normal_minus_one,
                     normal_zero,
                     normal_one) |> 
  as_tibble() |> 
  pivot_longer(cols = everything(),
               names_to = "distribution",
               values_to = "normal_draws")

normal_data |> 
  ggplot(aes(x = normal_draws)) +
  geom_histogram(color = "white", bins = 90) +
  facet_wrap(~ distribution, nrow = 3)
```

```{r}
normal_narrow <- rnorm(10000, sd = 0.5)
normal_standard <- rnorm(10000)
normal_wide <- rnorm(10000, sd = 1.5)

normal_data <- cbind(normal_narrow,
                     normal_standard,
                     normal_wide) |> 
  as_tibble() |> 
  pivot_longer(cols = everything(),
               names_to = "distribution",
               values_to = "normal_draws")

normal_data |> 
  ggplot(aes(x = normal_draws)) +
  geom_histogram(color = "white", bins = 90) +
  facet_wrap(~ distribution, nrow = 3)
```

### Sampling from Data

```{r}
county_elections <- read_csv(here("data", "county_elections.csv"))
```

```{r}
county_elections |> 
  sample_n(size = 10)
```

```{r}
county_elections |> 
  sample_frac(size = 0.01) # Return 1% of the final data
```

```{r}
county_elections |> 
  group_by(state_abbrv) |> 
  sample_n(1)
```

### set.seed()

```{r}
set.seed(10071992)

sample(my_animals, 3)
```

### Loops

```{r}
print("cat")
print("dog")
print("monkey")
print("parrot")
print("fish")
```

```{r}
my_animals <- c("cat", "dog", "monkey", "parrot", "fish")

for (animal in my_animals) {
  print(animal)
}
```

```{r}
# Make vector of integers from 1 to 10
1:10
```

```{r}
for (i in 1:10) {
  i <- i ^ 2
  print(i)
}
```

```{r}
holder <- NULL # Use NULL to create an empty object
for (i in 1:100) {
  holder[i] <- i ^ 2
}
```

```{r}
holder
```

```{r}
holder <- NULL
for (i in 1:100) {
  holder <- c(holder, i ^ 2)
}
```


### Simulations

### Central Limit Theorem

```{r}
n_sims <- 1000
means <- NULL
```

```{r}
for (i in 1:n_sims) {
  sample_data <- runif(100, min = -1, max = 1)
  means[i] <- mean(sample_data)
}
```

```{r}
tibble(means) |> 
  ggplot(aes(x = means)) +
  geom_histogram(color = "white", bins = 30) +
  labs(title = "Distribution of Means from Simulated Uniform Random Samples")
```

### Power Analysis

```{r}
n_sims <- 10000
p_values <- NULL
effect <- 0.5
sample_size <- 50
```

```{r}
# Setting seed so we all get the same results
set.seed(10071992)

for (i in 1:n_sims) {
  treated <- rnorm(sample_size / 2, mean = effect)
  control <- rnorm(sample_size / 2, mean = 0)
  
  power_test <- t.test(treated, control)
  p_values[i] <- power_test$p.value
}
```

```{r}
mean(p_values < 0.05)
```

### Bootstrapping

```{r}
n_sims <- 1000
estimates <- NULL
```

```{r}
for (i in 1:n_sims) {
  county_sample <- county_elections |> 
    sample_n(size = nrow(county_elections), replace = TRUE)
  
  lm_model <- lm(median_hh_inc ~ lesshs_pct,
                 data = county_sample)
  estimates[i] <- summary(lm_model)$coef[[2]]
}
```

```{r}
# saving ggplot object so we can add onto it later
bootstrap_plot <- tibble(estimates) |> 
  ggplot(aes(x = estimates)) +
  geom_histogram(color = "black",
                 fill = "white",
                 bins = 30) +
  labs(x = "Bootstrapped Estimates of the Regression Coefficient") +
  theme_bw()

bootstrap_plot
```

```{r}
lm_model <- lm(median_hh_inc ~ lesshs_pct,
               data = county_elections)
estimate <- summary(lm_model)$coef[[2]]
standard_error <- summary(lm_model)$coef[[4]]
```

```{r}
bootstrap_plot +
  geom_vline(xintercept = estimate,
             color = "maroon", size = 1.5) +
  geom_vline(xintercept = estimate + standard_error * 1.96,
             color = "darkcyan", size = 1.5, linetype = "dashed") +
  geom_vline(xintercept = estimate - standard_error * 1.96,
             color = "darkcyan", size = 1.5, linetype = "dashed") +
  labs(caption = "Full sample estimate in maroon\n95% confidence interval in cyan")
```