---
title: "Do Countries that Spend More on Public Health Have Better Health Outcomes?"
subtitle: "Example RMarkdown Report"
author: "Bertrand Wilden"
date: "`r Sys.Date()`"
output: pdf_document
header-includes:
   - \usepackage{floatrow}
   - \floatsetup[figure]{capposition=top}
bibliography: packages.bib
nocite: '@*'
---

```{r setup, include=FALSE}
# In this chunk I am setting some global options for all code chunks
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE)

# Loading in all necessary packages
# Don't forget to install.packages("package name") if you're missing any of these packages
library(dplyr)
library(ggplot2)
library(poliscidata)
library(texreg)

# This automatically generates a .bib file which contains the citation information for packages used
# the YAML 'bibliography: packages.bib' will automatically add the citations to the end of the document
knitr::write_bib(.packages(), "packages.bib")
```

```{r}
world_data <- world %>% 
  mutate(democracy = if_else(democ == "Yes", "Democracy", "Non-Democracy")) %>% 
  filter(!is.na(democracy))

lowest_spendhealth <- world_data %>% 
  filter(spendhealth == min(spendhealth))

highest_spendhealth <- world_data %>% 
  filter(spendhealth == max(spendhealth))
```

# Introduction

Around the world there is wide variation in how much countries spend on their public health services. At the low end is `r lowest_spendhealth$country`, which only spends `r lowest_spendhealth$spendhealth`% of its GPD on public health. On the other extreme is `r highest_spendhealth$country` whose public health expenditures are `r highest_spendhealth$spendhealth`% of its GPD. Figure 1 displays the full range of variation in health expenditures globally. But does how much a county spends on health services affect health outcomes, such as life expectancy, in that country?

```{r, fig.cap="The Distribution of Public Health Expenditures Globally"}
world_data %>% 
  ggplot(aes(x = spendhealth)) +
  geom_histogram(color = "white", fill = "cadetblue") +
  theme_minimal() +
  labs(x = "Public Expenditure on Health as a Percentage of GDP",
       y = "Frequency",
       caption = paste0("Mean: ", round(mean(world_data$spendhealth), 1),
                         "\nStandard Deviation: ", round(sd(world_data$spendhealth), 1))) +
  theme(text = element_text(family = "serif"))
```

# Analysis

We can test the strength of the relationship between public health expenditures and life expectancy by running the following bivariate regression model:

$$
\text{LifeExpectancy}_i = \alpha + \beta \text{HealthExpenditure}_i + \epsilon
$$

```{r}
model1 <- lm(lifeex_total ~ spendhealth, data = world_data)

model1_tex <- texreg(model1,
                     custom.model.names = "Life Expectancy",
                     custom.coef.names = c("(Intercept)", "Health Exp (% of GDP)"),
                     caption = "Bivariate OLS Regression")
```

`r model1_tex`

Table 1 shows the results of this simple model. As we can see, there is a positive and statistically significant relationship between public health expenditures and life expectancy in a country. For every extra percentage point of its GDP that a country spends on public health, we predict that the average life expectancy in that country will increase by `r round(coef(model1)[[2]], 2)` years. This strong positive association is also shown in Figure 2.

```{r, fig.cap="The Effect of Health Expenditure on Life Expectancy"}
p <- world_data %>% 
  ggplot(aes(x = spendhealth, y = lifeex_total, label = country)) +
  geom_text(check_overlap = TRUE, size = 3) +
  geom_smooth(method = "lm", color = "cadetblue", fill = "lightgray") +
  theme_minimal() +
  labs(x = "Public Expenditure on Health as a Percentage of GDP",
       y = "Life Expextancy at Birth") +
  theme(text = element_text(family = "serif"))
p
```

\newpage

# Software Packages
